Designing a high-performance boundary element library with OpenCL and Numba

### 2021-04-06 bem.tex

Once a triangulation is given we need to define the necessary data structures for the discretisation. Bempp-cl uses a Galerkin discretisation: we introduce set of basis functions \(\phi_1\) to \(\phi_N\), and define the \textbf{trial space} as the span of these function. We then approximate the solution \(\phi\) of \cref{eq:bnd_integral} by $\phi_h=\sum_{j=1}^N x_j\phi_j$. In the most simple case, we can define the function \(\phi_j\) to be equal to 1 on the triangle $\tau_j$ and 0 everywhere else. Other spaces are commonly defined to be piecewise polynomials on each triangle.

To discretise \cref{eq:bnd_integral}, we define a \textbf{test (or dual) space} in terms of a basis \(\psi_1\) to \(\psi_N\). The discrete representation of the above problem then takes the form
$$
Ax = b
$$
with
\begin{align*}
A_{ij} &= \int_{\Gamma}\psi_i(\bx)\int_{\Gamma}g(\bx, \by)\phi_j(\bx)\ds[\by]\ds[\bx]\\
b_i &= \int_{\Gamma}\psi_i(\bx)f(\bx)\ds[\bx].
\end{align*}
In the case of piecewise constant trial and test functions, the definition of $A_{ij}$ simplifies to $A_{ij} = \int_{\Gamma}\int_{\Gamma}g(\bx, \by)\ds[\by]\ds[\bx]$.

In Bempp-cl, an operator definition consists of the type of the operator (e.g. Laplace single-layer in the above example),
the definition of the trial and test spaces, and the definition of the range space. The range space is required for operator
products and not relevant for the purpose of this paper. The function $f$ is represented as a \textbf{grid function} object, which
consists of either the dual representation in the forms of the integrals $b_i = \int_{\Gamma}\psi_i(\bx)f(\bx)\ds[\bx]$ or directly through its coefficients $f_j$ in the representation $f=\sum_{j=1}^N f_j\phi_j$.

Once the grid, the space objects, and the operator(s) are defined, the main computational step is performed, namely the \textbf{computation of the discrete matrix entries} $A_{ij}$. For pairs of triangles $\tau_i$ and $\tau_j$ that do not share a joint edge or vertex this is done through a simple numerical quadrature rule that approximates $A_{ij}\approx \sum_{\ell}\sum_{q} g(\bx_\ell, \by_q)\psi_i(\bx_\ell)\phi_j(\by_q)\omega_\ell\omega_q$, where the $\bx_\ell$ and $\by_q$ are quadrature points in the corresponding triangles $\tau_i$ and $\tau_j$, and the values $\omega_i$ and $\omega_j$ are the quadrature weights. In the case that two triangles share a joint vertex/edge or the triangles $\tau_i$ and $\tau_j$ are identical, corresponding singularity adapted quadrature numerical quadrature rules are used that are based on singularity removing coordinate transformations.

The values $b_i$ of the right-hand side vector $b$ are similarly computed through a numerical quadrature rule.

In the final step, \textbf{Bempp solves the underlying linear system of equations} either through a direct LU decomposition or through iterative solvers such as GMRes. The solution can then be evaluated away from the surface $\Gamma$ through domain potential operators and exported in various formats for visualisation.

In summary, to solve a boundary integral equation problem, the following steps are performed by Bempp:
\begin{enumerate}
	\item Import of the surface description as triangulation data.
	\item Definition of the trial space, test space and range space.
	\item Discretisation into a matrix problem $Ax=b$.
	\item Solution of the matrix problem by either a direct or iterative solver.
	\item Evaluation of domain potential operators for visualisation and post-processing.
\end{enumerate}

>>>>

Once a triangulation is given we need to define the necessary data structures for the discretisation. Bempp-cl uses a Galerkin discretisation: we introduce set of basis functions \(\phi_1\) to \(\phi_N\), and define the \textbf{trial space} as the span of these function. We then approximate the solution \(\phi\) of \cref{eq:bnd_integral} by $\phi_h=\sum_{j=1}^N \dvec{x}_j\phi_j$, where $\dvec{x}$ is a vector of coefficients. In the most simple case, we can define the function \(\phi_j\) to be equal to 1 on the triangle $\tau_j$ and 0 everywhere else. Other spaces are commonly defined to be piecewise polynomials on each triangle.

To discretise \cref{eq:bnd_integral}, we define a \textbf{test (or dual) space} in terms of a basis \(\psi_1\) to \(\psi_N\). The discrete representation of the above problem then takes the form
$$
\dmat{A}\dvec{x} = \dvec{b}
$$
with
\begin{align*}
\dmat{A}_{ij} &= \int_{\Gamma}\psi_i(\bx)\int_{\Gamma}g(\bx, \by)\phi_j(\bx)\ds[\by]\ds[\bx]\\
\dvec{b}_i &= \int_{\Gamma}\psi_i(\bx)f(\bx)\ds[\bx].
\end{align*}
In the case of piecewise constant trial and test functions, the definition of $\dmat{A}_{ij}$ simplifies to $\dmat{A}_{ij} = \int_{\Gamma}\int_{\Gamma}g(\bx, \by)\ds[\by]\ds[\bx]$.

In Bempp-cl, an operator definition consists of the type of the operator (e.g. Laplace single-layer in the above example),
the definition of the trial and test spaces, and the definition of the range space. The range space is required for operator
products and not relevant for the purpose of this paper. The function $f$ is represented as a \textbf{grid function} object, which
consists of either the dual representation in the forms of the integrals $\dvec{b}_i = \int_{\Gamma}\psi_i(\bx)f(\bx)\ds[\bx]$ or directly through its coefficients $f_j$ in the representation $f=\sum_{j=1}^N f_j\phi_j$.

Once the grid, the space objects, and the operator(s) are defined, the main computational step is performed, namely the \textbf{computation of the discrete matrix entries} $\dmat{A}_{ij}$. For pairs of triangles $\tau_i$ and $\tau_j$ that do not share a joint edge or vertex this is done through a simple numerical quadrature rule that approximates $\dmat{A}_{ij}\approx \sum_{\ell}\sum_{q} g(\bx_\ell, \by_q)\psi_i(\bx_\ell)\phi_j(\by_q)\omega_\ell\omega_q$, where the $\bx_\ell$ and $\by_q$ are quadrature points in the corresponding triangles $\tau_i$ and $\tau_j$, and the values $\omega_i$ and $\omega_j$ are the quadrature weights. In the case that two triangles share a joint vertex/edge or the triangles $\tau_i$ and $\tau_j$ are identical, corresponding singularity adapted quadrature numerical quadrature rules are used that are based on singularity removing coordinate transformations.

The values $\dvec{b}_i$ of the right-hand side vector $b$ are similarly computed through a numerical quadrature rule.

In the final step, \textbf{Bempp solves the underlying linear system of equations} either through a direct LU decomposition or through iterative solvers such as GMRES. The solution can then be evaluated away from the surface $\Gamma$ through domain potential operators and exported in various formats for visualisation.

In summary, to solve a boundary integral equation problem, the following steps are performed by Bempp:
\begin{enumerate}
	\item Import of the surface description as triangulation data.
	\item Definition of the trial space, test space and range space.
	\item Discretisation into a matrix problem $\dmat{A}\dvec{x}=\dvec{b}$.
	\item Solution of the matrix problem by either a direct or iterative solver.
	\item Evaluation of domain potential operators for visualisation and post-processing.
\end{enumerate}

### 2021-04-06 bempp_overview.tex
add command for \ds

The main user-visible component of Bempp-cl is the module \texttt{bempp.api}, which defines all user interface functions and other high-level routines. In particular, it contains the definition of the main object types: \texttt{Grid}, \texttt{Space}, \texttt{GridFunction}, \texttt{Operator}. All computational backend routines are part of \texttt{bempp.core}. We are currently supporting a full Numba backend and an OpenCL backend. An overview of this structure is provided in \cref{fig:overview}.

Outside the operator discretisation, Numba is used in the following contexts:
\begin{itemize}
	\item Topological computations for the grid. Finding out all neighbour relationships between triangles requires parses through the grid data structure with linear complexity, which is accelerated with Numba.
	\item Definition of local-to-global maps for function spaces. Again, this requires traversal through the grid and assigning relationships between global and local indices.
	\item Grid Functions. A right-hand side function $f$ can be defined as a Python callable. This is just-in-time compiled via Numba and then the product with the corresponding basis functions integrated in each triangle via numerical quadrature, again via Numba accelerated routines.
	\item Sparse matrices. Sparse surface matrices are assembled through Numba accelerated routines.
\end{itemize}

The main computational complexity in discretising the boundary integral equation \cref{eq:bnd_integral} is coming from the discretisation of the left-hand side integral operator. Using dense methods this has quadratic complexity in the number of surface triangles.

Once the user defines an operator with its associated discretisation spaces and calls the \textbf{weak\_form} method to discretise it the code calls a regular integrator to assemble all the interactions between non-adjacent elements and a singular integrator to compute the interactions between adjacent triangles (only necessary if trial and test space are defined on the same grid). The result is stored internally as a dense Numpy array. The corresponding discretisation routines are proxies that forward to computational routines in either Numba or OpenCL, depending on the user preferences. For OpenCL assembly the code checks additional parameters, such as the default vector length for SIMD operators (e.g. 4 for double precision and 8 for single precision in Intel AVX2, or 1 if a GPU is used), and whether the discretisation should proceed in single or double precision. The OpenCL kernel is then compiled for the underlying compute device using PyOpenCL and executed. If the computational backend is Numba, the call is forwarded to the corresponding Numba discretisation routines and executed. For simple piecewise constant function spaces or other spaces, where the support of each basis function is localised to a single triangle, only one call to the computational routines is necessary. If the support of basis functions is larger than a single triangle different threads may need to sum into the same global degree of freedom. We discuss this in more detail in \cref{sec:coloring}.

>>>>

The main user-visible component of Bempp-cl is the module \pyth{bempp.api}, which defines all user interface functions and other high-level routines. In particular, it contains the definition of the main object types: \pyth{Grid}, \pyth{Space}, \pyth{GridFunction}, \pyth{Operator}. The computational backend routines are contained in the module \pyth{bempp.core}. Currently, we support a Numba and OpenCL backends. An overview of this structure is provided in \cref{fig:overview}.

The main computational cost involved in solving a problem using boundary element methods is due to discretising the boundary integral operator on the left-hand side of \cref{eq:bnd_integral} to obtain the matrix $\dmat{A}$: using dense methods, discretising an operator has quadratic complexity in terms of the number of surface triangles (although for larger problems this cost can be reduced through the use of hierarchical matrices or fast multipole methods). Great care is therefore taken to ensure that the operator assembly routines perform as highly as possible.

Once a user has defined in operator using Bempp-cl, the discretisation can be computed by calling the \pyth{weak_form} method. Upon calling this method, a regular integrator will be used to assemble all the interactions between non-adjacent elements, and a singular integrator will be used to compute the interactions between adjacent triangles (if the trial and test spaces are defined on different grids, this second integrator is not needed). Depending on the user's preferences, these integrators will internally use computational routines defined using either Numba or OpenCL.

For OpenCL assembly, the code checks additional parameters, such as the default vector length for SIMD operators (e.g., 4 for double precision and 8 for single precision in Intel AVX2, or 1 if a GPU is used), andwhether the discretisation should proceed in single or double precision. The OpenCL kernel is then compiled for the underlying compute device using PyOpenCL and executed. If the computational backend is Numba, the call is forwarded to the corresponding Numba discretisation routines and executed.

For simple piecewise constant function spaces or other spaces, where the support of each basis function is localised to a single triangle, only one call to the computational routines is necessary. If the support of basis functions is larger than a single triangle, different threads may need to sum into the same global degree of freedom. We discuss this in more detail in \cref{sec:coloring}.

Outside the operator discretisation, Numba is used in the following contexts:
\begin{itemize}
  \item Computing the grid topology: this involves iterating through the grid to compute the neighbour relataionships between triangles.
  \item Definition of local-to-global maps for function spaces: again, this requires traversal through the grid and assigning relationships between global and local indices.
  \item Grid functions: a right-hand side function $f$ can be defined as a Python callable. This is just-in-time compiled via Numba and then the product with the corresponding basis functions is integrated in each triangle via numerical quadrature, again via Numba accelerated routines.
  \item Computing sparse matrices.
\end{itemize}

As the cost of each of these processes is in general much smaller than the cost of operator discretisation, these can be permformed using Numba without any need to consider the use of OpenCL for potential further speed up.

### 2021-04-06 intro.tex

\chapterinitial{The Bempp boundary element library} (originally named BEM++ \cite{BEM++}) started in 2011 as a project to develop an open-source C++ library for the fast solution of boundary integral equations. The original release came with a simple Python wrapper to the C++ library. Over time, more and more functionality was moved into the Python interface, while computationally intensive routines and the main data structures remained in C++.

This proved a burden when efforts started to modernise the library to be able to make better use of Single-Instruction-Multiple-Data (SIMD) optimisation on CPUs and offload computation to GPUs. In 2018, we decided to rewrite the computational core from scratch. The aims were to support explicit SIMD optimisation on CPUs with various instruction lengths, be able to offload computations to AMD, Intel, and Nvidia GPUs, and to base the complete codebase on Python. These aims naturally led to the choice of building a Python library based around OpenCL (using the PyOpenCL interface) and Numba.

>>>>

\chapterinitial{The Bempp boundary element library} (originally named BEM++) started in 2011 as a project to develop an open-source C++ library for the fast solution of boundary integral equations. The original release came with a simple Python wrapper to the C++ library. Over time, more and more functionality was moved into the Python interface, while computationally intensive routines and the main data structures remained in C++.

This proved a burden when efforts started to modernise the library to be able to make better use of single-instruction-multiple-data (SIMD) optimisation on CPUs and offload computation to GPUs. In 2018, we decided to rewrite the computational core from scratch. The aims were to support explicit SIMD optimisation on CPUs with various instruction lengths, be able to offload computations to AMD, Intel, and Nvidia GPUs, and to base the complete codebase on Python. These aims naturally led to the choice of building a Python library based around OpenCL (using the PyOpenCL interface) and Numba.

### 2021-04-06 opencl.tex

The above description was valid for kernels that do not take advantage of CPU SIMD operations are kernels that run directly on a GPU device. For SIMD optimization on CPUs we proceed slightly differently. The corresponding kernel in principle works as described above, but batches together the interaction on one test triangle with X trial triangles, where X is either 4, 8, or 16, depending on the number of available SIMD lanes. This strategy allows us to optimize almost all floating point operations within a kernel run for SIMD operation. If the number of trial elements is not divisible by 4, 8, or 16, then the few remaining trial elements are assembled with the standard-non vectorized kernel.

Each kernel definition is stored in two variants, one with the ending \textit{\_novec.cl} and another one with the ending \textit{\_vec.cl}. The vectorized variant is configured via preprocessor directives for the desired number of vector lanes. Having to develop two OpenCL kernel codes for each operator is a certain amount of overhead. If we write a new kernel, we first implement the non-vectorized version. Then, with the help of preprocessor directives and a number of helper functions that do the actual implementation of operations depending on whether the kernel is vectorized or not, it is usually a matter of an hour or two to convert the non-vectorized kernel into a vectorized version.

Alternatively, some CPU OpenCL runtime environments have the option to try and auto-vectorize kernels by batching together work items on SIMD lanes, similarly to what we do manually. In our experience this works well for very simple kernels but often fails for more complex OpenCL kernels. This is why we decided to implement this strategy manually.

>>>>

The above description was valid for kernels that do not take advantage of CPU SIMD operations are kernels that run directly on a GPU device. For SIMD optimisation on CPUs we proceed slightly differently. The corresponding kernel in principle works as described above, but batches together the interaction on one test triangle with X trial triangles, where X is either 4, 8, or 16, depending on the number of available SIMD lanes. This strategy allows us to optimise almost all floating point operations within a kernel run for SIMD operation. If the number of trial elements is not divisible by 4, 8, or 16, then the few remaining trial elements are assembled with the standard-non vectorised kernel.

Each kernel definition is stored in two variants, one with the ending \textit{\_novec.cl} and another one with the ending \textit{\_vec.cl}. The vectorised variant is configured via preprocessor directives for the desired number of vector lanes. Having to develop two OpenCL kernel codes for each operator is a certain amount of overhead. If we write a new kernel, we first implement the non-vectorised version. Then, with the help of preprocessor directives and a number of helper functions that do the actual implementation of operations depending on whether the kernel is vectorised or not, it is usually a matter of an hour or two to convert the non-vectorised kernel into a vectorised version.

Alternatively, some CPU OpenCL runtime environments have the option to try and auto-vectorise kernels by batching together work items on SIMD lanes, similarly to what we do manually. In our experience this works well for very simple kernels but often fails for more complex OpenCL kernels. This is why we decided to implement this strategy manually.
